---
title: "R Notebook"
output: html_notebook
---

```{r Load Packages}
library(limma)
library(minfi)
library(IlluminaHumanMethylation450kanno.ilmn12.hg19)
library(IlluminaHumanMethylation450kmanifest)
library(RColorBrewer)
library(missMethyl)
library(matrixStats)
library(minfiData)
library(Gviz)
library(DMRcate)
library(stringr)
library(sva)
```

### Import Data and Annotations ###

```{r}
ann450k = getAnnotation(IlluminaHumanMethylation450kanno.ilmn12.hg19)

dataDirectory <- "/Users/kendallanderson/Documents/HMA Project/MethylSeq/methylation/iDaT_files"

targets <- read.metharray.sheet(dataDirectory, pattern="SampleSheet.csv")
targets

# create methyl array object
rgSet <- read.metharray.exp(targets=targets)
rgSet
# assign the samples names to the rgSet
sampleNames(rgSet) <- targets$Sample_Name
rgSet
```

### Quality Control ###

```{r Quality Control}
# calculate detection pvalues based on control probed in 450k array
detP <- detectionP(rgSet)
head(detP)

# plot p-values
pal <- brewer.pal(8,"Dark2")
par(mfrow=c(1,2))
barplot(colMeans(detP), col=pal[factor(targets$Cell_status)], las=2,
 cex.names=0.6,ylab="Mean detection p-values")
abline(h=0.01,col="red")
legend("topleft", legend=levels(factor(targets$Cell_status)), fill=pal,
 bg="white")
barplot(colMeans(detP), col=pal[factor(targets$Cell_status)], las=2,
 cex.names=0.6, ylim = c(0,0.002), ylab="Mean detection p-values")
legend("topleft", legend=levels(factor(targets$Cell_status)), fill=pal,
 bg="white")
```
*No low quality samples so no need to throw any out.*

### Normalization ###

```{r Normalization}
library(harrypotter)
pal <- hp(n = 4, house = "Gryffindor")

# test SWAN and quantile normalization
mSetSWAN <- preprocessSWAN(rgSet)
mSetQuan <- preprocessQuantile(rgSet)
# raw dataset, has to be converted into same object type
mSetRaw <- preprocessRaw(rgSet)

mSetNorm <- preprocessFunnorm(rgSet)

# plot swan normalization vs raw
par(mfrow=c(1,2))
densityPlot(rgSet, sampGroups=targets$Resistance_status,main="Raw", legend=FALSE)
legend("top", legend = levels(factor(targets$Resistance_status)),
 text.col=brewer.pal(8,"Dark2"))
densityPlot(getBeta(mSetSWAN), sampGroups=targets$Resistance_status,
 main="SWAN Normalized", legend=FALSE)
legend("top", legend = levels(factor(targets$Resistance_status)),
 text.col=brewer.pal(8,"Dark2"))

# plot quantile normalization vs raw
tiff("Meth_Normalization.tiff", units="in", width=7, height=5, res=300)
par(mfrow=c(1,2))
densityPlot(rgSet, sampGroups=targets$Status,main="Raw", legend=FALSE, pal=hp(n=4, option="lunalovegood"))
legend("top", legend = levels(factor(targets$Status)),
 text.col=hp(n=4, option="lunalovegood"))
densityPlot(getBeta(mSetQuan), sampGroups=targets$Status,
 main="Quantile Normalized", legend=FALSE, pal=hp(n=4, option="lunalovegood"))
legend("top", legend = levels(factor(targets$Status)),
 text.col=hp(n=4, option="lunalovegood"))
dev.off()

# plot funnorm normalization vs raw
par(mfrow=c(1,2))
densityPlot(rgSet, sampGroups=targets$Status,main="Raw", legend=FALSE, pal=hp(n=4, option="lunalovegood"))
legend("top", legend = levels(factor(targets$Status)),
 text.col=hp(n=4, option="lunalovegood"))
densityPlot(getBeta(mSetNorm), sampGroups=targets$Status,
 main="Quan Normalized", legend=FALSE, pal=hp(n=4, option="lunalovegood"))
legend("top", legend = levels(factor(targets$Status)),
 text.col=hp(n=4, option="lunalovegood"))
```
*Quantile normalization seems to yield better results.*

### Data Filtering ###

```{r Filter Probes}
mSetQuanFlt <- dropLociWithSnps(mSetQuan)
mSetQuanFlt
```

### Batch Correction ###

```{r Test Different Batch Correction Options}
# perform different options on beta
cell_batch_corr_beta <- ComBat(getBeta(mSetQuanFlt), batch=targets$Sample_Group, mod = model.matrix(~as.factor(Cell_status), data=targets))

res_batch_corr_beta <- ComBat(getBeta(mSetQuanFlt), batch=targets$Sample_Group, mod = model.matrix(~as.factor(Resistance_status), data=targets))

treat_batch_corr_beta <- ComBat(getBeta(mSetQuanFlt), batch=targets$Sample_Group, mod = model.matrix(~as.factor(Treatment_status), data=targets))

unsup_batch_corr_beta <- ComBat(getBeta(mSetQuanFlt), batch=targets$Sample_Group)

# perform different options on M
cell_batch_corr_M <- ComBat(getM(mSetQuanFlt), batch=targets$Sample_Group, mod = model.matrix(~as.factor(Cell_status), data=targets))

res_batch_corr_M <- ComBat(getM(mSetQuanFlt), batch=targets$Sample_Group, mod = model.matrix(~as.factor(Resistance_status), data=targets))

treat_batch_corr_M <- ComBat(getM(mSetQuanFlt), batch=targets$Sample_Group, mod = model.matrix(~as.factor(Treatment_status), data=targets))

unsup_batch_corr_M <- ComBat(getM(mSetQuanFlt), batch=targets$Sample_Group)
```

```{r Plot M Batch Correction Comparisons}
par(mfrow=c(1,2))
densityPlot(getM(mSetQuanFlt), sampGroups=targets$Resistance_status,
 main="Quan Normalized M", legend=FALSE)
legend("bottom", legend = levels(factor(targets$Resistance_status)),
 text.col=brewer.pal(8,"Dark2"))
densityPlot(cell_batch_corr_M, sampGroups=targets$Resistance_status,
 main="Batch Corrected- Cell M", legend=FALSE)
legend("bottom", legend = levels(factor(targets$Resistance_status)),
 text.col=brewer.pal(8,"Dark2"))

par(mfrow=c(1,2))
densityPlot(getM(mSetQuanFlt), sampGroups=targets$Resistance_status,
 main="Quan Normalized M", legend=FALSE)
legend("bottom", legend = levels(factor(targets$Resistance_status)),
 text.col=brewer.pal(8,"Dark2"))
densityPlot(res_batch_corr_M, sampGroups=targets$Resistance_status,
 main="Batch Corrected- Res M", legend=FALSE)
legend("bottom", legend = levels(factor(targets$Resistance_status)),
 text.col=brewer.pal(8,"Dark2"))

par(mfrow=c(1,2))
densityPlot(getM(mSetQuanFlt), sampGroups=targets$Resistance_status,
 main="Quan Normalized M", legend=FALSE)
legend("bottom", legend = levels(factor(targets$Resistance_status)),
 text.col=brewer.pal(8,"Dark2"))
densityPlot(treat_batch_corr_M, sampGroups=targets$Resistance_status,
 main="Batch Corrected- Treat M", legend=FALSE)
legend("bottom", legend = levels(factor(targets$Resistance_status)),
 text.col=brewer.pal(8,"Dark2"))

par(mfrow=c(1,2))
densityPlot(getM(mSetQuanFlt), sampGroups=targets$Resistance_status,
 main="Quan Normalized M", legend=FALSE)
legend("bottom", legend = levels(factor(targets$Resistance_status)),
 text.col=brewer.pal(8,"Dark2"))
densityPlot(unsup_batch_corr_M, sampGroups=targets$Resistance_status,
 main="Batch Corrected- Unsup M", legend=FALSE)
legend("bottom", legend = levels(factor(targets$Resistance_status)),
 text.col=brewer.pal(8,"Dark2"))
```

```{r Plot Beta Batch Correction Comparison}
par(mfrow=c(1,2))
densityPlot(getBeta(mSetQuanFlt), sampGroups=targets$Resistance_status,
 main="Quan Normalized", legend=FALSE)
legend("top", legend = levels(factor(targets$Resistance_status)),
 text.col=brewer.pal(8,"Dark2"))
densityPlot(cell_batch_corr, sampGroups=targets$Resistance_status,
 main="Batch Corrected- Cell", legend=FALSE)
legend("top", legend = levels(factor(targets$Resistance_status)),
 text.col=brewer.pal(8,"Dark2"))

par(mfrow=c(1,2))
densityPlot(getBeta(mSetQuanFlt), sampGroups=targets$Resistance_status,
 main="Quan Normalized", legend=FALSE)
legend("top", legend = levels(factor(targets$Resistance_status)),
 text.col=brewer.pal(8,"Dark2"))
densityPlot(res_batch_corr, sampGroups=targets$Resistance_status,
 main="Batch Corrected- Res", legend=FALSE)
legend("top", legend = levels(factor(targets$Resistance_status)),
 text.col=brewer.pal(8,"Dark2"))

par(mfrow=c(1,2))
densityPlot(getBeta(mSetQuanFlt), sampGroups=targets$Resistance_status,
 main="Quan Normalized", legend=FALSE)
legend("top", legend = levels(factor(targets$Resistance_status)),
 text.col=brewer.pal(8,"Dark2"))
densityPlot(treat_batch_corr, sampGroups=targets$Resistance_status,
 main="Batch Corrected- Treat", legend=FALSE)
legend("top", legend = levels(factor(targets$Resistance_status)),
 text.col=brewer.pal(8,"Dark2"))

tiff("Meth_Batch_Correction.tiff", units="in", width=7, height=5, res=300)
par(mfrow=c(1,2))
densityPlot(getBeta(mSetQuanFlt), sampGroups=targets$Status,
 main="Normalized Pre-Correction", legend=FALSE, pal=hp(n=4, option="lunalovegood"))
legend("top", legend = levels(factor(targets$Status)),
 text.col=hp(n=4, option="lunalovegood"))
densityPlot(unsup_batch_corr, sampGroups=targets$Status,
 main="Normalized Batch Corrected", legend=FALSE, pal=hp(n=4, option="lunalovegood"))
legend("top", legend = levels(factor(targets$Status)),
 text.col=hp(n=4, option="lunalovegood"))
dev.off()
```
*Decided to go with the unsupervised batch correction.*

```{r Plot all three (raw, norm, batch corr)}
png("Meth_Batch_Correction.png", units="in", width=10, height=5, res=300)
par(mfrow=c(1,3))
densityPlot(rgSet, sampGroups=targets$Status,main="Raw", legend=FALSE, pal=hp(n=4, option="lunalovegood"))
legend("top", legend = levels(factor(targets$Status)),
 text.col=hp(n=4, option="lunalovegood"))
densityPlot(getBeta(mSetQuanFlt), sampGroups=targets$Status,
 main="Normalized Pre-Correction", legend=FALSE, pal=hp(n=4, option="lunalovegood"))
legend("top", legend = levels(factor(targets$Status)),
 text.col=hp(n=4, option="lunalovegood"))
densityPlot(unsup_batch_corr, sampGroups=targets$Status,
 main="Normalized Batch Corrected", legend=FALSE, pal=hp(n=4, option="lunalovegood"))
legend("top", legend = levels(factor(targets$Status)),
 text.col=hp(n=4, option="lunalovegood"))
dev.off()
```

### MDS Plots ###

```{r MDS Plot Before Batch Correction}
par(mfrow=c(1,3))
plotMDS(getM(mSetQuanFlt), top=1000, gene.selection="common",
 col=pal[factor(targets$Cell_status)])
legend("top", legend=levels(factor(targets$Cell_status)), text.col=pal,
 bg="white", cex=0.7)
plotMDS(getM(mSetQuanFlt), top=1000, gene.selection="common",
 col=pal[factor(targets$Resistance_status)])
legend("top", legend=levels(factor(targets$Resistance_status)), text.col=pal,
 bg="white", cex=0.7)
plotMDS(getM(mSetQuanFlt), top=1000, gene.selection="common",
 col=pal[factor(targets$Resistance_status)])
legend("top", legend=levels(factor(targets$Treatment_status)), text.col=pal,
 bg="white", cex=0.7)
```

```{r MDS Plot After Batch Correction}
par(mfrow=c(1,3))
plotMDS(unsup_batch_corr_M, top=1000, gene.selection="common",
 col=pal[factor(targets$Cell_status)])
legend("top", legend=levels(factor(targets$Cell_status)), text.col=pal,
 bg="white", cex=0.7)
plotMDS(unsup_batch_corr_M, top=1000, gene.selection="common",
 col=pal[factor(targets$Resistance_status)])
legend("top", legend=levels(factor(targets$Resistance_status)), text.col=pal,
 bg="white", cex=0.7)
plotMDS(unsup_batch_corr_M, top=1000, gene.selection="common",
 col=pal[factor(targets$Resistance_status)])
legend("top", legend=levels(factor(targets$Treatment_status)), text.col=pal,
 bg="white", cex=0.7)
```

```{r MDS Plot for Betas After Cell Batch Correction}
par(mfrow=c(1,3))
plotMDS(cell_batch_corr_beta, top=1000, gene.selection="common",
 col=pal[factor(targets$Cell_status)])
legend("top", legend=levels(factor(targets$Cell_status)), text.col=pal,
 bg="white", cex=0.7)
plotMDS(cell_batch_corr_beta, top=1000, gene.selection="common",
 col=pal[factor(targets$Resistance_status)])
legend("top", legend=levels(factor(targets$Resistance_status)), text.col=pal,
 bg="white", cex=0.7)
plotMDS(cell_batch_corr_beta, top=1000, gene.selection="common",
 col=pal[factor(targets$Resistance_status)])
legend("top", legend=levels(factor(targets$Treatment_status)), text.col=pal,
 bg="white", cex=0.7)
```

```{r}
deltas <- data.frame("R1"=(unsup_batch_corr_beta[,2])-(unsup_batch_corr_beta[,1]),
                     "R2"=(unsup_batch_corr_beta[,6])-(unsup_batch_corr_beta[,5]),
                     "R3"=(unsup_batch_corr_beta[,3])-(unsup_batch_corr_beta[,8]),
                     "R4"=(unsup_batch_corr_beta[,7])-(unsup_batch_corr_beta[,4]),
                     "T1"=(unsup_batch_corr_beta[,8])-(unsup_batch_corr_beta[,1]),
                     "T2"=(unsup_batch_corr_beta[,4])-(unsup_batch_corr_beta[,5]),
                     "T3"=(unsup_batch_corr_beta[,3])-(unsup_batch_corr_beta[,2]),
                     "T4"=(unsup_batch_corr_beta[,7])-(unsup_batch_corr_beta[,6]))

avg_deltas <- data.frame("R1_avg"=mean(deltas[,1]),
                         "R2_avg"=mean(deltas[,2]),
                         "R3_avg"=mean(deltas[,3]),
                         "R4_avg"=mean(deltas[,4]),
                         "T1_avg"=mean(deltas[,5]),
                         "T2_avg"=mean(deltas[,6]),
                         "T3_avg"=mean(deltas[,7]),
                         "T4_avg"=mean(deltas[,8]))
# annotate all betas
betas_annotated <- as.data.frame(cbind2(ann450k[match(rownames(deltas),ann450k$Name),c(1:4,24,26)],deltas))

```

```{r Gene Lists for Sig Deltas- Resistance Comparisons}
# select for sig deltas for R1
R1_deltas_filt <- subset(deltas, abs(R1)>0.3, select="R1")
# annotate with gene names
R1_sig_genes <- as.data.frame(cbind2(ann450k[match(rownames(R1_deltas_filt),ann450k$Name),c(1:4,24,26)],R1_deltas_filt))
# remove any position that does not have an associated gene
R1_sig_genes <- subset(R1_sig_genes, UCSC_RefGene_Name!="")
# change UCSC annotations to character strings
R1_sig_genes$UCSC_RefGene_Group <- as.character(R1_sig_genes$UCSC_RefGene_Group)
R1_sig_genes$UCSC_RefGene_Name <- as.character(R1_sig_genes$UCSC_RefGene_Name)
# split gene lists and position lists
R1_name <- strsplit(R1_sig_genes$UCSC_RefGene_Name, split=";")
R1_group <- strsplit(R1_sig_genes$UCSC_RefGene_Group, split=";")
# create data frame with gene and pos matching deltas
R1_sig_genes <- data.frame(R1_delta = rep(R1_sig_genes$R1, sapply(R1_name, length)), 
                           GeneID = unlist(R1_name), 
                           cgPos = unlist(R1_group),
                           chr = rep(R1_sig_genes$chr, sapply(R1_name, length)),
                           pos = rep(R1_sig_genes$pos, sapply(R1_name, length)),
                           strand = rep(R1_sig_genes$strand, sapply(R1_name, length)))
# remove duplicates
R1_sig_genes <- unique(R1_sig_genes)
# write out file
#write.csv(R1_sig_genes, file = "/Users/kendallanderson/Documents/HMA Project/MethylSeq/methylation/Gene_Lists/R1_sig_genes.csv")

# select for sig deltas for R2
R2_deltas_filt <- subset(deltas, abs(R2)>0.3, select="R2")
# annotate with gene names
R2_sig_genes <- as.data.frame(cbind2(ann450k[match(rownames(R2_deltas_filt),ann450k$Name),c(1:4,24,26)],R2_deltas_filt))
# remove any position that does not have an associated gene
R2_sig_genes <- subset(R2_sig_genes, UCSC_RefGene_Name!="")
# change UCSC annotations to character strings
R2_sig_genes$UCSC_RefGene_Group <- as.character(R2_sig_genes$UCSC_RefGene_Group)
R2_sig_genes$UCSC_RefGene_Name <- as.character(R2_sig_genes$UCSC_RefGene_Name)
# split gene lists and position lists
R2_name <- strsplit(R2_sig_genes$UCSC_RefGene_Name, split=";")
R2_group <- strsplit(R2_sig_genes$UCSC_RefGene_Group, split=";")
# create data frame with gene and pos matching deltas
R2_sig_genes <- data.frame(R2_delta = rep(R2_sig_genes$R2, sapply(R2_name, length)), 
                           GeneID = unlist(R2_name), 
                           cgPos = unlist(R2_group),
                           chr = rep(R2_sig_genes$chr, sapply(R2_name, length)),
                           pos = rep(R2_sig_genes$pos, sapply(R2_name, length)),
                           strand = rep(R2_sig_genes$strand, sapply(R2_name, length)))
# remove duplicates
R2_sig_genes <- unique(R2_sig_genes)
# write out file
#write.csv(R2_sig_genes, file = "/Users/kendallanderson/Documents/HMA Project/MethylSeq/methylation/Gene_Lists/R2_sig_genes.csv")

# select for sig deltas for R3
R3_deltas_filt <- subset(deltas, abs(R3)>0.3, select="R3")
# annotate with gene names
R3_sig_genes <- as.data.frame(cbind2(ann450k[match(rownames(R3_deltas_filt),ann450k$Name),c(1:4,24,26)],R3_deltas_filt))
# remove any position that does not have an associated gene
R3_sig_genes <- subset(R3_sig_genes, UCSC_RefGene_Name!="")
# change UCSC annotations to character strings
R3_sig_genes$UCSC_RefGene_Group <- as.character(R3_sig_genes$UCSC_RefGene_Group)
R3_sig_genes$UCSC_RefGene_Name <- as.character(R3_sig_genes$UCSC_RefGene_Name)
# split gene lists and position lists
R3_name <- strsplit(R3_sig_genes$UCSC_RefGene_Name, split=";")
R3_group <- strsplit(R3_sig_genes$UCSC_RefGene_Group, split=";")
# create data frame with gene and pos matching deltas
R3_sig_genes <- data.frame(R3_delta = rep(R3_sig_genes$R3, sapply(R3_name, length)), 
                           GeneID = unlist(R3_name), 
                           cgPos = unlist(R3_group),
                           chr = rep(R3_sig_genes$chr, sapply(R3_name, length)),
                           pos = rep(R3_sig_genes$pos, sapply(R3_name, length)),
                           strand = rep(R3_sig_genes$strand, sapply(R3_name, length)))
# remove duplicates
R3_sig_genes <- unique(R3_sig_genes)
# write out file
#write.csv(R3_sig_genes, file = "/Users/kendallanderson/Documents/HMA Project/MethylSeq/methylation/Gene_Lists/R3_sig_genes.csv")

# select for sig deltas for R4
R4_deltas_filt <- subset(deltas, abs(R4)>0.3, select="R4")
# annotate with gene names
R4_sig_genes <- as.data.frame(cbind2(ann450k[match(rownames(R4_deltas_filt),ann450k$Name),c(1:4,24,26)],R4_deltas_filt))
# remove any position that does not have an associated gene
R4_sig_genes <- subset(R4_sig_genes, UCSC_RefGene_Name!="")
# change UCSC annotations to character strings
R4_sig_genes$UCSC_RefGene_Group <- as.character(R4_sig_genes$UCSC_RefGene_Group)
R4_sig_genes$UCSC_RefGene_Name <- as.character(R4_sig_genes$UCSC_RefGene_Name)
# split gene lists and position lists
R4_name <- strsplit(R4_sig_genes$UCSC_RefGene_Name, split=";")
R4_group <- strsplit(R4_sig_genes$UCSC_RefGene_Group, split=";")
# create data frame with gene and pos matching deltas
R4_sig_genes <- data.frame(R4_delta = rep(R4_sig_genes$R4, sapply(R4_name, length)), 
                           GeneID = unlist(R4_name), 
                           cgPos = unlist(R4_group),
                           chr = rep(R4_sig_genes$chr, sapply(R4_name, length)),
                           pos = rep(R4_sig_genes$pos, sapply(R4_name, length)),
                           strand = rep(R4_sig_genes$strand, sapply(R4_name, length)))
# remove duplicates
R4_sig_genes <- unique(R4_sig_genes)
# write out file
#write.csv(R4_sig_genes, file = "/Users/kendallanderson/Documents/HMA Project/MethylSeq/methylation/Gene_Lists/R4_sig_genes.csv")
```

```{r Gene Lists for Sig Deltas- Treatment Comparisons}
# select for sig deltas for T1
T1_deltas_filt <- subset(deltas, abs(T1)>0.3, select="T1")
# annotate with gene names
T1_sig_genes <- as.data.frame(cbind2(ann450k[match(rownames(T1_deltas_filt),ann450k$Name),c(1:4,24,26)],T1_deltas_filt))
# remove any position that does not have an associated gene
T1_sig_genes <- subset(T1_sig_genes, UCSC_RefGene_Name!="")
# change UCSC annotations to character strings
T1_sig_genes$UCSC_RefGene_Group <- as.character(T1_sig_genes$UCSC_RefGene_Group)
T1_sig_genes$UCSC_RefGene_Name <- as.character(T1_sig_genes$UCSC_RefGene_Name)
# split gene lists and position lists
T1_name <- strsplit(T1_sig_genes$UCSC_RefGene_Name, split=";")
T1_group <- strsplit(T1_sig_genes$UCSC_RefGene_Group, split=";")
# create data frame with gene and pos matching deltas
T1_sig_genes <- data.frame(T1_delta = rep(T1_sig_genes$T1, sapply(T1_name, length)), 
                           GeneID = unlist(T1_name), 
                           cgPos = unlist(T1_group),
                           chr = rep(T1_sig_genes$chr, sapply(T1_name, length)),
                           pos = rep(T1_sig_genes$pos, sapply(T1_name, length)),
                           strand = rep(T1_sig_genes$strand, sapply(T1_name, length)))
# remove duplicates
T1_sig_genes <- unique(T1_sig_genes)
# write out file
#write.csv(T1_sig_genes, file = "/Users/kendallanderson/Documents/HMA Project/MethylSeq/methylation/Gene_Lists/T1_sig_genes.csv")

# select for sig deltas for T2
T2_deltas_filt <- subset(deltas, abs(T2)>0.3, select="T2")
# annotate with gene names
T2_sig_genes <- as.data.frame(cbind2(ann450k[match(rownames(T2_deltas_filt),ann450k$Name),c(1:4,24,26)],T2_deltas_filt))
# remove any position that does not have an associated gene
T2_sig_genes <- subset(T2_sig_genes, UCSC_RefGene_Name!="")
# change UCSC annotations to character strings
T2_sig_genes$UCSC_RefGene_Group <- as.character(T2_sig_genes$UCSC_RefGene_Group)
T2_sig_genes$UCSC_RefGene_Name <- as.character(T2_sig_genes$UCSC_RefGene_Name)
# split gene lists and position lists
T2_name <- strsplit(T2_sig_genes$UCSC_RefGene_Name, split=";")
T2_group <- strsplit(T2_sig_genes$UCSC_RefGene_Group, split=";")
# create data frame with gene and pos matching deltas
T2_sig_genes <- data.frame(T2_delta = rep(T2_sig_genes$T2, sapply(T2_name, length)), 
                           GeneID = unlist(T2_name), 
                           cgPos = unlist(T2_group),
                           chr = rep(T2_sig_genes$chr, sapply(T2_name, length)),
                           pos = rep(T2_sig_genes$pos, sapply(T2_name, length)),
                           strand = rep(T2_sig_genes$strand, sapply(T2_name, length)))
# remove duplicates
T2_sig_genes <- unique(T2_sig_genes)
# write out file
#write.csv(T2_sig_genes, file = "/Users/kendallanderson/Documents/HMA Project/MethylSeq/methylation/Gene_Lists/T2_sig_genes.csv")

# select for sig deltas for T3
T3_deltas_filt <- subset(deltas, abs(T3)>0.3, select="T3")
# annotate with gene names
T3_sig_genes <- as.data.frame(cbind2(ann450k[match(rownames(T3_deltas_filt),ann450k$Name),c(1:4,24,26)],T3_deltas_filt))
# remove any position that does not have an associated gene
T3_sig_genes <- subset(T3_sig_genes, UCSC_RefGene_Name!="")
# change UCSC annotations to character strings
T3_sig_genes$UCSC_RefGene_Group <- as.character(T3_sig_genes$UCSC_RefGene_Group)
T3_sig_genes$UCSC_RefGene_Name <- as.character(T3_sig_genes$UCSC_RefGene_Name)
# split gene lists and position lists
T3_name <- strsplit(T3_sig_genes$UCSC_RefGene_Name, split=";")
T3_group <- strsplit(T3_sig_genes$UCSC_RefGene_Group, split=";")
# create data frame with gene and pos matching deltas
T3_sig_genes <- data.frame(T3_delta = rep(T3_sig_genes$T3, sapply(T3_name, length)), 
                           GeneID = unlist(T3_name), 
                           cgPos = unlist(T3_group),
                           chr = rep(T3_sig_genes$chr, sapply(T3_name, length)),
                           pos = rep(T3_sig_genes$pos, sapply(T3_name, length)),
                           strand = rep(T3_sig_genes$strand, sapply(T3_name, length)))
# remove duplicates
T3_sig_genes <- unique(T3_sig_genes)
# write out file
#write.csv(T3_sig_genes, file = "/Users/kendallanderson/Documents/HMA Project/MethylSeq/methylation/Gene_Lists/T3_sig_genes.csv")

# select for sig deltas for T4
T4_deltas_filt <- subset(deltas, abs(T4)>0.3, select="T4")
# annotate with gene names
T4_sig_genes <- as.data.frame(cbind2(ann450k[match(rownames(T4_deltas_filt),ann450k$Name),c(1:4,24,26)],T4_deltas_filt))
# remove any position that does not have an associated gene
T4_sig_genes <- subset(T4_sig_genes, UCSC_RefGene_Name!="")
# change UCSC annotations to character strings
T4_sig_genes$UCSC_RefGene_Group <- as.character(T4_sig_genes$UCSC_RefGene_Group)
T4_sig_genes$UCSC_RefGene_Name <- as.character(T4_sig_genes$UCSC_RefGene_Name)
# split gene lists and position lists
T4_name <- strsplit(T4_sig_genes$UCSC_RefGene_Name, split=";")
T4_group <- strsplit(T4_sig_genes$UCSC_RefGene_Group, split=";")
# create data frame with gene and pos matching deltas
T4_sig_genes <- data.frame(T4_delta = rep(T4_sig_genes$T4, sapply(T4_name, length)), 
                           GeneID = unlist(T4_name), 
                           cgPos = unlist(T4_group),
                           chr = rep(T4_sig_genes$chr, sapply(T4_name, length)),
                           pos = rep(T4_sig_genes$pos, sapply(T4_name, length)),
                           strand = rep(T4_sig_genes$strand, sapply(T4_name, length)))
# remove duplicates
T4_sig_genes <- unique(T4_sig_genes)
# write out file
#write.csv(T4_sig_genes, file = "/Users/kendallanderson/Documents/HMA Project/MethylSeq/methylation/Gene_Lists/T4_sig_genes.csv")
```

```{r Write out unique gene lists}
R1_genes <- unique(R1_sig_genes$GeneID)
R2_genes <- unique(R2_sig_genes$GeneID)
R3_genes <- unique(R3_sig_genes$GeneID)
R4_genes <- unique(R4_sig_genes$GeneID)
T1_genes <- unique(T1_sig_genes$GeneID)
T2_genes <- unique(T2_sig_genes$GeneID)
T3_genes <- unique(T3_sig_genes$GeneID)
T4_genes <- unique(T4_sig_genes$GeneID)

#write.csv(R1_genes, file="/Users/kendallanderson/Documents/HMA Project/MethylSeq/methylation/Gene_Lists/R1_gene_list.csv")
#write.csv(R2_genes, file="/Users/kendallanderson/Documents/HMA Project/MethylSeq/methylation/Gene_Lists/R2_gene_list.csv")
#write.csv(R3_genes, file="/Users/kendallanderson/Documents/HMA Project/MethylSeq/methylation/Gene_Lists/R3_gene_list.csv")
#write.csv(R4_genes, file="/Users/kendallanderson/Documents/HMA Project/MethylSeq/methylation/Gene_Lists/R4_gene_list.csv")
#write.csv(T1_genes, file="/Users/kendallanderson/Documents/HMA Project/MethylSeq/methylation/Gene_Lists/T1_gene_list.csv")
#write.csv(T2_genes, file="/Users/kendallanderson/Documents/HMA Project/MethylSeq/methylation/Gene_Lists/T2_gene_list.csv")
#write.csv(T3_genes, file="/Users/kendallanderson/Documents/HMA Project/MethylSeq/methylation/Gene_Lists/T3_gene_list.csv")
#write.csv(T4_genes, file="/Users/kendallanderson/Documents/HMA Project/MethylSeq/methylation/Gene_Lists/T4_gene_list.csv")
```


```{r}
nrow(R1_sig_genes)
length(R1_genes)

nrow(subset(T4_sig_genes, T4_delta>0))
nrow(subset(T4_sig_genes, T4_delta<0))
```

```{r Histograms of Delta Distributions}
hist(deltas$R1, breaks=50, xlim=c(-1,1), main="Differential Betas in Resistant Cells", xlab="Delta Values", ylim=c(0,180000))
hist(deltas$R2, breaks=50, xlim=c(-1,1), main="Differential Betas in Resistant Cells", xlab="Delta Values", ylim=c(0,180000))

hist(deltas$R3, breaks=50, xlim=c(-1,1))
hist(deltas$R4, breaks=50, xlim=c(-1,1))

hist(deltas$T1, breaks=50, xlim=c(-1,1), main="Differential Beta Values in Treated Cells", xlab="Delta Values", ylim=c(0,180000))
hist(deltas$T2, breaks=50, xlim=c(-1,1), main="Differential Beta Values in Treated Cells", xlab="Delta Values", ylim=c(0,180000))
hist(deltas$T3, breaks=50, xlim=c(-1,1), main="Differential Beta Values in Treated Resistant Cells", xlab="Delta Values", ylim=c(0,180000))
hist(deltas$T4, breaks=50, xlim=c(-1,1), main="Differential Betas in Treated Resistant Cells", xlab="Delta Values", ylim=c(0,180000))
```

```{r Overlapping histograms of delta values}
# make dataframe to generate overlapping histograms
deltas_one <- data.frame("Delta"=c(deltas$R1,deltas$R2,deltas$R3,deltas$R4,deltas$T1,deltas$T2,deltas$T3,deltas$T4),"Comparison"=c(rep("R1",nrow(deltas)),rep("R2",nrow(deltas)),rep("R3",nrow(deltas)),rep("R4",nrow(deltas)),rep("T1",nrow(deltas)),rep("T2",nrow(deltas)),rep("T3",nrow(deltas)),rep("T4",nrow(deltas))))

png("Meth_R1R2_hist.png", units="in", width=7, height=5, res=300)
# plot deltas for parental vs resistant
ggplot(deltas_one,aes(x=Delta)) + 
    geom_histogram(data=subset(deltas_one,Comparison == 'R1'),fill = "#8fb9ba", alpha = 0.6, binwidth = 0.01) +
    geom_histogram(data=subset(deltas_one,Comparison == 'R2'),fill = "#9a8fba", alpha = 0.6, binwidth = 0.01) +
  xlim(c(-0.75,0.75)) +
  #labs(title="Changes in Methylation Upon Gaining Resistance") + 
  ylab("Frequency") + xlab("Delta Values") + geom_vline(xintercept=c(-0.3,0.3), linetype="dotted")
dev.off()

png("Meth_T1T2_hist.png", units="in", width=7, height=5, res=300)
# plot deltas for untreated vs treated
ggplot(deltas_one,aes(x=Delta)) + 
    geom_histogram(data=subset(deltas_one,Comparison == 'T1'),fill = "#8fb9ba", alpha = 0.6, binwidth = 0.01) +
    geom_histogram(data=subset(deltas_one,Comparison == 'T2'),fill = "#9a8fba", alpha = 0.6, binwidth = 0.01) +
  xlim(c(-0.75,0.75)) + 
  #labs(title="Distribution of Delta Values for Treated Cells") + 
  ylab("Frequency") + xlab("Delta Values") +
  geom_vline(xintercept=c(-0.3,0.3), linetype="dotted")
dev.off()

png("Meth_T3T4_hist.png", units="in", width=7, height=5, res=300)
# plot deltas for untreated vs treated, resistant
ggplot(deltas_one,aes(x=Delta)) + 
    geom_histogram(data=subset(deltas_one,Comparison == 'T3'),fill = "#8fb9ba", alpha = 0.6, binwidth = 0.01) +
    geom_histogram(data=subset(deltas_one,Comparison == 'T4'),fill = "#9a8fba", alpha = 0.6, binwidth = 0.01) +
  xlim(c(-0.75,0.75)) +
  #labs(title="Distribution of Delta Values for Treated Resistant Cells") + 
  ylab("Frequency") + xlab("Delta Values") + geom_vline(xintercept=c(-0.3,0.3), linetype="dotted")
dev.off()
```

```{r}
nrow(subset(deltas, T4>0.3))
nrow(subset(deltas, T4<(-0.3)))
```

```{r Annotating the Deltas}
ann450kSub <- ann450k[match(rownames(deltas),ann450k$Name),
 c(1:4,12:19,24:ncol(ann450k))]
```

```{r}
densityBeanPlot(unsup_batch_corr_beta, sampGroups = targets$Cell_status)
```

```{r}
par(mfrow=c(1,2))
densityPlot(unsup_batch_corr_beta, sampGroups=targets$Resistance_status, main="Beta values",
 legend=FALSE, xlab="Beta values")
legend("top", legend = levels(factor(targets$Resistance_status)),
 text.col=brewer.pal(8,"Dark2"))
densityPlot(unsup_batch_corr_M, sampGroups=targets$Resistance_status, main="M-values",
 legend=FALSE, xlab="M values")
legend("topleft", legend = levels(factor(targets$Resistance_status)),
 text.col=brewer.pal(8,"Dark2"))
```

### Probe-Wise Differential Methylation Analysis ###

## This is not working and I am very frustrated. ##
```{r}
# this is the factor of interest
status <- factor(targets$Status)
# this is the individual effect that we need to account for
cell <- factor(targets$Cell_status)

# use the above to create a design matrix
design <- model.matrix(~0+status+cell, data=targets)
colnames(design) <- c(levels(status),levels(cell)[-1])

# fit the linear model
fit <- lmFit(unsup_batch_corr_M, design)
# create a contrast matrix for specific comparisons
contMatrix <- makeContrasts(Par_UT-Res_UT,
                            Par_T-Res_T,
                            Par_UT-Par_T,
                            Res_UT-Res_T,
                            levels=design)
contMatrix

#fit2 <- contrasts.fit(fit, contMatrix)
#fit2 <- eBayes(fit2)

#summary(decideTests(fit2))
cell
```

```{r}
dmp <- dmpFinder(unsup_batch_corr_beta, pheno=targets$Status, type="categorical")
dmp <- cbind(dmp, ID=rownames(dmp))
dmp_annot_combined <- cbind(ann450k[row.names(dmp),],dmp)
```


```{r}
View(dmp_annot_combined)
```


```{r Res1}
dmp_res1 <- dmpFinder(unsup_batch_corr_beta[,1:2], pheno=targets$Status)
dmp_res1 <- cbind(dmp_res1, ID=rownames(dmp_res1))
dmp_res1_annot_combined <- cbind(ann450k[row.names(dmp_res1),],dmp_res1)
```

~~~

### Annotate Genes of Interest from RNA HEL Overlaps ###

```{r}
View(HEL_deltas)
# subset HEL and overlaps of interest
HEL_deltas <- deltas[,c(2,6,8)]
# annotate with gene names
HEL_deltas <- as.data.frame(cbind2(ann450k[match(rownames(HEL_deltas),ann450k$Name),c(1:4,24,26)],HEL_deltas))
# remove any position that does not have an associated gene
HEL_deltas <- subset(HEL_deltas, UCSC_RefGene_Name!="")
# change UCSC annotations to character strings
HEL_deltas$UCSC_RefGene_Group <- as.character(HEL_deltas$UCSC_RefGene_Group)
HEL_deltas$UCSC_RefGene_Name <- as.character(HEL_deltas$UCSC_RefGene_Name)
# split gene lists and position lists
HEL_name <- strsplit(HEL_deltas$UCSC_RefGene_Name, split=";")
HEL_group <- strsplit(HEL_deltas$UCSC_RefGene_Group, split=";")
# create data frame with gene and pos matching deltas
HEL_deltas_final <- data.frame(R2_delta = rep(HEL_deltas$R2, sapply(HEL_name, length)),
                               T2_delta = rep(HEL_deltas$T2, sapply(HEL_name, length)),
                               T4_delta = rep(HEL_deltas$T4, sapply(HEL_name, length)),
                               GeneID = unlist(HEL_name), 
                               cgPos = unlist(HEL_group),
                               chr = rep(HEL_deltas$chr, sapply(HEL_name, length)),
                               pos = rep(HEL_deltas$pos, sapply(HEL_name, length)),
                               strand = rep(HEL_deltas$strand, sapply(HEL_name, length)))

# remove duplicates
HEL_deltas_final <- unique(HEL_deltas_final)

# select for genes of interest
HEL_deltas_four <- subset(HEL_deltas_final, GeneID=="ATF3" | GeneID=="DDIT4" | GeneID=="RMRP" | GeneID=="RPPH1")

HEL_deltas_twelve <- subset(HEL_deltas_final, GeneID=="AGAP1" | GeneID=="AK4" | GeneID=="ASNS" | GeneID=="C5orf4" | GeneID=="CHAC1" | GeneID=="EGR3" | GeneID=="FAM83A" | GeneID=="GNG8" | GeneID=="HBZ" | GeneID=="INHA" | GeneID=="LOC100505633" | GeneID=="RPTOR")

# write out file
#write.csv(HEL_deltas_four, file = "/Users/kendallanderson/Documents/HMA Project/MethylSeq/methylation/Gene_Lists/HEL_deltas_four.csv")
#write.csv(HEL_deltas_twelve, file = "/Users/kendallanderson/Documents/HMA Project/MethylSeq/methylation/Gene_Lists/HEL_deltas_twelve.csv")
```


```{r}
HEL_deltas_two_opp <- subset(HEL_deltas_final, GeneID=="BMPER" | GeneID=="CDH3")
write.csv(HEL_deltas_two_opp, file="/Users/kendallanderson/Documents/HMA Project/MethylSeq/methylation/Gene_Lists/HEL_deltas_two_opp.csv")

View(HEL_deltas_two_opp)
```

```{r Plot Methylation for BMPER and CDH3}
# plot BMPER methylation
ggplot(subset(HEL_deltas_two_opp, GeneID=="BMPER"), aes(x=T2_delta, y=T4_delta)) +
  geom_point(aes(col=cgPos), size=2, alpha=0.7) +
  xlim(-0.7,0.7) +
  ylim(-0.7,0.7) +
  ggtitle("HEL Treatment Comparison- BMPER") + xlab("Sensitive Delta") + ylab("Resistant Delta") +
  geom_hline(yintercept=0, alpha=0.5) +
  geom_vline(xintercept=0, alpha=0.5) +
  geom_vline(xintercept=c(-0.3,0.3), linetype="dotted", alpha=0.5) +
  geom_hline(yintercept=c(-0.3,0.3), linetype="dotted", alpha=0.5) +
  scale_color_manual(labels = c("1stExon", "3'UTR", "Body", "TSS1500", "TSS200"),
                    values=c("1stExon"="#202547",
                             "3'UTR"="#53354A",
                             "Body"="#BE3144",
                             "TSS1500"="#FF7844",
                             "TSS200"="#E3D26F"),
                    name="Position") 

# plot CDH3 methylation
ggplot(subset(HEL_deltas_two_opp, GeneID=="CDH3"), aes(x=T2_delta, y=T4_delta)) +
  geom_point(aes(col=cgPos), size=2, alpha=0.7) +
  xlim(-0.7,0.7) +
  ylim(-0.7,0.7) +
  ggtitle("HEL Treatment Comparison- CDH3") + xlab("Sensitive Delta") + ylab("Resistant Delta") +
  geom_hline(yintercept=0, alpha=0.5) +
  geom_vline(xintercept=0, alpha=0.5) +
  geom_vline(xintercept=c(-0.3,0.3), linetype="dotted", alpha=0.5) +
  geom_hline(yintercept=c(-0.3,0.3), linetype="dotted", alpha=0.5) +
  scale_color_manual(labels = c("1stExon", "3'UTR", "5'UTR", "Body", "TSS1500", "TSS200"),
                    values=c("1stExon"="#202547",
                             "3'UTR"="#53354A",
                             "5'UTR"="#7A6C5D",
                             "Body"="#BE3144",
                             "TSS1500"="#FF7844",
                             "TSS200"="#E3D26F"),
                    name="Position") 

```






